This is a minimal framework of rules for experienced players of fantasy
roleplaying games. It assumes you know that ``3d6'' means to sum three
rolls of a six-sided die. It assumes you know that one player takes the
role of impartial Referee, who prepares an adventure in the form of a
keyed map full of monsters, strange wonders, and treasures. It assumes
you know that the remaining players each take the role of an
inexperienced but ambitious Player Character.

Furthermore, these rules assume players want enough creative room for
each campaign to grow through play into something unique from the last
campaign.

\section{Characters}\label{characters}

For each ability, roll 3d6 and record the total:

\begin{itemize}
\tightlist
\item
  Strength
\item
  Intelligence
\item
  Wisdom
\item
  Dexterity
\item
  Constitution
\item
  Charisma
\end{itemize}

After rolling all six ability scores, the player may swap two of them if
desired.

\subsection{Ability Score Bonuses}\label{ability-score-bonuses}

\begin{longtable}[]{@{}ll@{}}
\toprule
Score & Modifier\tabularnewline
\midrule
\endhead
3 & -3\tabularnewline
4--5 & -2\tabularnewline
6--8 & -1\tabularnewline
9--12 & 0\tabularnewline
13--15 & +1\tabularnewline
16--17 & +2\tabularnewline
18 & +3\tabularnewline
\bottomrule
\end{longtable}

\subsection{Character Advancement}\label{character-advancement}

\begin{longtable}[]{@{}ll@{}}
\toprule
Level & Experince Points\tabularnewline
\midrule
\endhead
1 & 0\tabularnewline
2 & 400\tabularnewline
3 & 1,200\tabularnewline
4 & 3,600\tabularnewline
5 & 10,000\tabularnewline
6 & 32,000\tabularnewline
7 & 64,000\tabularnewline
8 & 130,000\tabularnewline
9 & 260,000\tabularnewline
10 & 389,000\tabularnewline
11 & 500,000\tabularnewline
12 & 600,000\tabularnewline
13 & 700,000\tabularnewline
14 & 800,000\tabularnewline
15 & 900,000\tabularnewline
16 & 1,000,000\tabularnewline
\bottomrule
\end{longtable}

From first through ninth level, characters gain an additional hit die
each level, plus any constitution modifier. At level ten and above,
characters gain only two additional hit points per level. All hit dice
are d6. However, fighters roll two six-sided hit dice, and choose the
higher roll.

Characters gain at most one level per session. Reduce the experience
award of a character who would gain two levels so that their XP totals
just shy of the second level.

\subsection{Number of Daily Spells per Spell
Level}\label{number-of-daily-spells-per-spell-level}

\begin{longtable}[]{@{}lllllll@{}}
\toprule
Character Level & 1 & 2 & 3 & 4 & 5 & 6\tabularnewline
\midrule
\endhead
1 & 1 & & & & &\tabularnewline
2 & 2 & & & & &\tabularnewline
3 & 2 & 1 & & & &\tabularnewline
4 & 3 & 2 & & & &\tabularnewline
5 & 3 & 2 & 1 & & &\tabularnewline
6 & 4 & 2 & 2 & & &\tabularnewline
7 & 4 & 3 & 2 & 1 & &\tabularnewline
8 & 4 & 3 & 3 & 2 & &\tabularnewline
9 & 4 & 3 & 3 & 2 & 1 &\tabularnewline
10 & 4 & 4 & 3 & 2 & 2 &\tabularnewline
11 & 4 & 4 & 3 & 3 & 2 & 1\tabularnewline
12 & 4 & 4 & 4 & 3 & 2 & 1\tabularnewline
13 & 5 & 5 & 4 & 4 & 4 & 2\tabularnewline
14 & 5 & 5 & 4 & 4 & 4 & 3\tabularnewline
15 & 5 & 5 & 5 & 4 & 4 & 4\tabularnewline
16 & 5 & 5 & 5 & 5 & 5 & 5\tabularnewline
\bottomrule
\end{longtable}

Clerics do not gain additional spells after level 10.

\subsection{Clerics}\label{clerics}

Clerics are militant religious zealots and hunters of the undead.

Each day, they pray for particular spells from their spell list. At any
point during play, a cleric may swap out a prepared spell for the Turn
Undead spell.

Clerics use any armor, but only blunt weapons.

At ninth level, clerics establish a religious order of their own,
attract devotees, and collect tithes.

\subsection{Fighters}\label{fighters}

Fighters use any weapons or armor.

At ninth level, fighters may become lords of their own domain, clearing
wilderness to build a castle, raising an army, and levying taxes.

\subsection{Magic-Users}\label{magic-users}

Magic-users wear no armor, and lack expertise in any weapon demanding
more skill or strength than a simple dagger.

Each day, magic-users wrestle incantations from the pages of spell books
into their brains. A cast spell escapes the magic-user's mind
completely, unable to be cast again without further sleep and study.

\subsection{Attack Bonus}\label{attack-bonus}

All characters start at first level with an attack bonus of +1, which
they add to all their d20 attack rolls. Fighters gain an additional +1
per level (e.g.~a fourth-level fighter gets +4). Other classes gain an
additional plus one at levels three, six, nine, twelve, and fifteen. A
sixth-level cleric, for example, gets an attack bonus of +3.

Monsters use an attack bonus equal to their hit dice.

\subsection{Hirelings and Henchmen}\label{hirelings-and-henchmen}

Henchmen and hirelings work for the player characters, although they are
non-player characters controlled by the referee.

Hirelings are normal humans with a single hit die. They work for a fixed
rate, do not share in treasure, and gain no experience. They tolerate
only limited risks, commensurate with their pay. Hirelings include
guards, link-boys, bearers, sages, smiths, animal trainers, engineers,
alchemists, and sailors.

Henchmen have personal loyalty to a player character. They work for a
half-share of found treasure. Accordingly, they have a characters class,
and gain experience. Henchmen tolerate greater risks than hirelings (but
not abuse). Each henchman has a loyalty score (2d6), modified by the
player character's charisma. Extreme situations or demands may trigger a
loyalty check by the referee. A henchman who fails a loyalty check might
not return for the next delve, might run away immediately, or might even
turn against the player character --- biding time until the best
opportunity for betrayal.

\begin{longtable}[]{@{}lll@{}}
\toprule
Charisma & Maximum Henchmen & Loyalty Modifier\tabularnewline
\midrule
\endhead
3 & 1 & -3\tabularnewline
4--5 & 2 & -2\tabularnewline
6--8 & 3 & -1\tabularnewline
9--12 & 4 & 0\tabularnewline
13--15 & 5 & +1\tabularnewline
16--17 & 6 & +2\tabularnewline
18 & 12 & +3\tabularnewline
\bottomrule
\end{longtable}

\subsection{Random Dungeon Room
Stocking}\label{random-dungeon-room-stocking}

After thoughtfully placing unique monsters and treasures, the referee
may stock the remainder of a large dungeon area with the following
table.

\begin{longtable}[]{@{}ll@{}}
\toprule
d12 & Room Contents\tabularnewline
\midrule
\endhead
1--3 & Empty\tabularnewline
4 & Trick/trap (no treasure)\tabularnewline
5--7 & Monster (no treasure)\tabularnewline
8--10 & Treasure \& Monster\tabularnewline
11 & Treasure \& trick/trap\tabularnewline
12 & Treasure (concealed/hidden)\tabularnewline
\bottomrule
\end{longtable}

\subsection{Magic-User Spells Level 1}\label{magic-user-spells-level-1}

\begin{enumerate}
\def\labelenumi{\arabic{enumi}.}
\tightlist
\item
  Charm Person
\item
  Detect Magic
\item
  Hold Portal
\item
  Light
\item
  Read Magic
\item
  Sleep
\end{enumerate}

\subsection{Magic-User Spells Level 2}\label{magic-user-spells-level-2}

\begin{enumerate}
\def\labelenumi{\arabic{enumi}.}
\tightlist
\item
  Detect Invisibility
\item
  ESP
\item
  Invisibility
\item
  Knock
\item
  Locate Object
\item
  Phantasmal Force
\end{enumerate}

\subsection{Magic-User Spells Level 3}\label{magic-user-spells-level-3}

\begin{enumerate}
\def\labelenumi{\arabic{enumi}.}
\tightlist
\item
  Clairvoyance
\item
  Dispel Magic
\item
  Fire Ball
\item
  Fly
\item
  Lightning Bolt
\item
  Water Breathing
\end{enumerate}

\subsection{Magic-User Spells Level 4}\label{magic-user-spells-level-4}

\begin{enumerate}
\def\labelenumi{\arabic{enumi}.}
\tightlist
\item
  Charm Monster
\item
  Confusion
\item
  Dimension Door
\item
  Polymorph Others
\item
  Polymorph Self
\item
  Wizard Eye
\end{enumerate}

\subsection{Magic-User Spells Level 5}\label{magic-user-spells-level-5}

\begin{enumerate}
\def\labelenumi{\arabic{enumi}.}
\tightlist
\item
  Animate Dead
\item
  Conjure Elemental
\item
  Contact Higher Plane
\item
  Feeblemind
\item
  Magic Jar
\item
  Teleport
\end{enumerate}

\subsection{Magic-User Spells Level 6}\label{magic-user-spells-level-6}

\begin{enumerate}
\def\labelenumi{\arabic{enumi}.}
\tightlist
\item
  Control Weather
\item
  Disintegrate
\item
  Geas
\item
  Invisible Stalker
\item
  Reincarnation
\item
  Stone/Flesh
\end{enumerate}

\subsection{Cleric Spells Level 1}\label{cleric-spells-level-1}

\begin{enumerate}
\def\labelenumi{\arabic{enumi}.}
\tightlist
\item
  Cure Light Wounds*
\item
  Detect Evil*
\item
  Protection from Evil*
\item
  Purify Food \& Water*
\end{enumerate}

\subsection{Cleric Spells Level 2}\label{cleric-spells-level-2}

\begin{enumerate}
\def\labelenumi{\arabic{enumi}.}
\tightlist
\item
  Bless*
\item
  Hold Person
\item
  Silence 15' Radius
\item
  Speak with Animals
\end{enumerate}

\subsection{Cleric Spells Level 3}\label{cleric-spells-level-3}

\begin{enumerate}
\def\labelenumi{\arabic{enumi}.}
\tightlist
\item
  Cure Disease*
\item
  Growth of Animals
\item
  Remove Curse
\item
  Speak with Dead
\end{enumerate}

\subsection{Cleric Spells Level 4}\label{cleric-spells-level-4}

\begin{enumerate}
\def\labelenumi{\arabic{enumi}.}
\tightlist
\item
  Cure Serious Wounds*
\item
  Neutralize Poison
\item
  Protection from Evil 10' radius
\item
  Speak with Plants
\end{enumerate}

\subsection{Cleric Spells Level 5}\label{cleric-spells-level-5}

\begin{enumerate}
\def\labelenumi{\arabic{enumi}.}
\tightlist
\item
  Commune
\item
  Dispel Evil*
\item
  Insect Plague
\item
  Raise Dead*
\end{enumerate}

\section{Referee}\label{referee}

\subsection{Monsters}\label{monsters}

Suitable monsters include creatures from mythology, giant or mutant
animals, the undead, fairytale beings, antagonistic human beings,
carnivorous plants, fungi, dinosaurs, and even sci-fi elements, like
robots or Frankenstein's monster. Referees can expand the roster of
monsters by enhancing existing ones (fire-breathing goblin, basilisk
with bat wings) or creating new monsters from scratch (use the table
below as a starting point for the numbers).

four rounds times d6 damage = 3.5 times 4 = 14

\begin{longtable}[]{@{}llllll@{}}
\toprule
HD* & AC & Save & Dmg/Rnd** & XP & Treasure Value (GP)\tabularnewline
\midrule
\endhead
\textless{}1 & 9 {[}10{]} & 16 & d3 (2) & 5 & 20\tabularnewline
1 & 7 {[}11{]} & 15 & d6 (3.5) & 15 & 60\tabularnewline
2 & 7 {[}12{]} & 14 & d6 (3.5) & 30 & 120\tabularnewline
3 & 6 {[}13{]} & 13 & d6+1 or d8 (4.5) & 60 & 240\tabularnewline
4 & 6 {[}13{]} & 12 & d6+1 or d8 (4.5) & 120 & 480\tabularnewline
5 & 5 {[}14{]} & 11 & d6+2 or d10 (5.5) & 240 & 960\tabularnewline
6 & 5 {[}14{]} & 10 & d6+2 or d10 (5.5) & 400 & 1600\tabularnewline
7 & 4 {[}15{]} & 9 & d12 (6.5) & 600 & 2400\tabularnewline
8 & 4 {[}15{]} & 8 & d12 (6.5) & 800 & 3200\tabularnewline
9 & 3 {[}16{]} & 7 & 7.5) & 1100 & 4400\tabularnewline
10 & 3 {[}16{]} & 6 & 8.5) & 1400 & 5600\tabularnewline
11 & 2 {[}17{]} & 5 & 8.5) & 1700 & 6800\tabularnewline
12 & 2 {[}17{]} & 4 & 9.5) & 2000 & 8000\tabularnewline
13 & 1 {[}18{]} & 4 & 9.5) & 2300 & 9200\tabularnewline
14 & 1 {[}18{]} & 4 & 10.5 & 2600 & 10400\tabularnewline
15 & 0 {[}19{]} & 4 & 10.5 & 2900 & 11600\tabularnewline
16 & 0 {[}19{]} & 4 & & 3100 & 12400\tabularnewline
\bottomrule
\end{longtable}

\begin{itemize}
\tightlist
\item
  The treasure and XP award may match the values of a higher hit die
  monster, if the monster has more than one or two special abilities.
\end{itemize}

** Average per round damage over the first three or four rounds. A 6 HD
monster might do d6+2 damage every round, or we might instead have it do
d6 most rounds with a once-time d6+6 special attack.

Baalroch Demon AC: 2 {[}17{]} HD: 9 (HDE 10) Move: 6 (15 flying) XP:
1400 Wreathed in flame, these bull-like, bat-winged demons resist magic
75\%, and casters below 6th level affect them not at all. They wield a
sword and whip (used to pull enemies into their flames for 3d6 damage).
Their m\^{}\{e\}l'\{e\}e attack deals 2d6 damage.

Basilisk AC: 4 {[}15{]} HD: 6 (HDE 8) Move: 6 XP: 800 The gaze of these
eight-legged reptiles turns living things to stone.

Black Pudding AC: 6 {[}13{]} HD: 10 (HDE 11) Move: 6 XP: 1700 This
amoeba-like scavenger flows over ceilings as easily as floors, and
squeezes though openings as small as 1``. Its excretions dissolve wood
and metal. It's immune to cold, and split/multiplied by lightening.

Bugbear AC: 5 {[}14{]} HD: 3+1 (HDE 3) Move: 9 XP: 120 Awkward-looking
yet stealthy, bands of these hairy goblinoids surprise opponents half
the time.

Cockatrice AC: 6 {[}13{]} HD: 5 (HDE 7) Move: 6 (18 flying) XP: 600
Those bitten by a cockatrice turn to stone, unless they make a saving
throw.

Dragon

Djinn AC: 5 {[}14{]} HD: 7+1 (HDE 9) Move: 9 (24 flying) XP: 1100 Djinni
create food and water, and cloth or wooden objects. They become
invisible or assume gaseous form at will, or turn into a 10' whirlwind
that sweeps away creatures of 1 HD or less. Some grant wishes. Their
m\^{}\{e\}l'\{e\}e attack deals 2d6 damage.

Doppelganger AC: 5 {[}14{]} HD: 4 (HDE 5) Move: 9 XP: 240 Doppelgangers
transform to look like anyone. They're immune to Sleep and Charm, and
save +5 versus all magic.

Elemental

Eye of Terror AC: 0 (body) 2 (central eye) 3 (eye stalks) {[}19, 17,
16{]} HD: 10 (HDE 14) Move: 6 levitating XP: 2600 Its tentacles are
poison (save or die). Each round, it attacks with its central eye and
three eye stalks:

\begin{longtable}[]{@{}ll@{}}
\toprule
Eye & Power\tabularnewline
\midrule
\endhead
Central & Flesh to stone ray\tabularnewline
1 & Alter time (slow)\tabularnewline
2 & Polymorph\tabularnewline
3 & Hold monster\tabularnewline
4 & Sleep\tabularnewline
5 & Hold person\tabularnewline
6 & Stun ray\tabularnewline
7 & Telekinesis\tabularnewline
8 & Feeblemind\tabularnewline
\bottomrule
\end{longtable}

Gargoyle Ghoul Giant Goblin

Gray Ooze AC: 8 {[}11{]} HD: 3 (HDE 5) Move: 1 XP: 240 It's immune to
spells, blunt weapons, and any heat/cold damage. Any metal object it
contacts is dissolved (save vs.~acid).

Gelatinous Cube AC: 8 {[}11{]} HD: 4 (HDE 5) Move: 6 XP: 240 Hit
opponents are anesthetized (save vs.~paralysis), then enveloped in 6
rounds. Lightning and cold don't harm gelatinous cubes.

Green Slime Any metal or organic object touching it turns to green slime
(save allowed), though Cure Disease halts the transformation. Fire or
cold kills green slime.

Hippogriff Hobgoblin Hydra Invisible Stalker Kobold

Liche AC: 0 {[}19{]} HD: 12--18 (HDE 15--21) Move: 6 XP: 2900, 3200,
3500, 3800, 4100, 4400, 4700 They're undead wizards, with magic-user
levels equal to their hit dice. Their touch paralyzes (no save); even
seeing a liche paralyzes creatures with 4 or fewer HD.

Manticore Medusa

Mimic AC 7 {[}12{]} HD: 7--10 (HDE 8--11) Move: 3 XP: 800, 1100, 1400,
1700 Mimics take the shape of any wood, stone, or metal object, such as
a treasure chest.

Minotaur Mummy

Ocher Jelly AC: 8 {[}11{]} HD: 5 (HDE 6) Move: 3 XP: 400 Their acid
dissolves killed opponents (making Raise Dead nearly impossible).

Ogre AC: 5 {[}14{]} HD: 4+1 (HDE 4) Move: 9 XP: 120 They enslave
captives (1--2 on d8) or simply eat them (3--8). Orgres themselves are
sometimes enslaved by demons.

Orc AC: 6 {[}13{]} HD: 1 (HDE 1) Move: 12 XP: 15 They fight at -1 to hit
in sunlight. Leaders have more hit dice, and some tribes have a
magic-using shaman.

Owlbear AC 5 {[}14{]} HD: 5 (HDE 5) Move: 12 XP: 240

Pegasus Skeleton Specter Troll Vampire Werewolf Wight Wraith

Yellow Mold AC: N/A HD: N/A (HDE 3) Move: 0 XP: 60 If cut or punctured,
it releases a 10' diameter cloud of poisonous spores (save or die).
Touching yellow mold caused d6 acid damage. Fire destroys it.

Zombie

Rocs, Sumatran Rats, and Giant Serpents And dungeon vermin? Giant rat,
giant spider, giant flea, giant snail
